---
title: "Assignment 4 - Heart rate, respiration and interpersonal coordination"
author: "Gustav, Magnus, Morten, Nicoline, Tobias"
date: ""
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
pacman::p_load(tidyverse, ggplot2, dplyr, groupdata2, naniar, lmerTest)
```

# Assignment 4 - Heart rate, respiration and interpersonal coordination

In this assignment, you will learn to:
- pre-process physiological data (and grow further your mad R skills)
- model the continuous interdependence between two signals (using a multilevel model as proxy for a dynamical system approach)
- conservatively assess the presence of coordination between to signals in a controlled context

This assignment has two parts. 
1. The first part familiarizes you with heart rate, and respiration data and their preprocessing. 
2. The second part explores how to analyze interpersonal coordination of these signals.

These are the questions you need to be able to answer at the end of the assignment (aka that you need to submit as part of the portfolio)

1) How do you preprocess heart rate and respiration data? Describe the process. If any data needs to be excluded, list the excluded data and motivate the exclusion.

2) Do you observe interpersonal coordination in heart rate and respiration? Describe your control baseline, the method used to quantify coordination, and the statistical models used to infer whether coordination was higher than in the baseline. Report the results of the models.

3) Do you observe differences in coordination between conditions? Report the models and results.

4) Is respiration coordination a likely driver of heart rate coordination? Describe how you would test for it. Bonus points if you actually run the tests and report methods and results.

N.B. N.B. The data are collected by students from previous years (Study 1 - 4). Note that synchronous and turn-taking are the same across all four studies, but the third condition is different: in the first year it was self-paced joint reading; in the second to fourth years it was the tv-series conversation.

## Let's get started

### Exploring physiological signals
The data files can be found here: https://www.dropbox.com/sh/bvvk7t3fvsplh9o/AADM6q4WrtXKvSwH5aAO1umta?dl=0

- Choose one pair (one pair, three conditions, three files)
- Load the logs
- Produce a plot of the participants' respiration signal and a different one of the participants' HR signal.
  N.B: remember the slides: artifacts, downsampling, scaling.
  N.B. The gridExtra::grid.arrange() function allows you to display the plots side by side. E.g. grid.arrange(plot1, plot2, plot3, ncol=3). There are also smarter packages, like cowplot and ggpubr.
- Can you eye-ball which condition if any displays more physiological coordination?

### First we read one data file and identify the procedure
- Load the file
- correctly identify all columns
- plot the data
- deal with the artifacts
- downsample the data
- Add a column for study, group, trial and condition

```{r}
# Load the libraries

# Load the file
data <- read_csv("data/Study2_G5_T1_Synchronous.csv")
data$row_number <- seq.int(nrow(data))
str(data)

# Plot
data %>% ggplot(aes(x = time)) + 
  geom_line(aes(y = Resp1, color = 'Resp1')) + 
  geom_line(aes(y = Resp2, color = 'Resp2')) +
  labs(title = "Respiration") +
  labs(x = "Time", y = "Respiration") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, size = 8),  
        panel.grid.minor = element_blank())

data %>% ggplot(aes(x = time)) + 
  geom_line(aes(y = HR1, color = 'HR1')) + 
  geom_line(aes(y = HR2, color = 'HR2')) + 
  labs(title = "Heart Rate") +
  labs(x = "Time", y = "Heart Rate") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, size = 8), 
        panel.grid.minor = element_blank())  

## Remove outliers ###################


# Plot raw data against those with the artifacts removed
#Heart Rate plots 
gridExtra::grid.arrange(data %>% ggplot(aes(x = time, y = HR1)) + geom_line() + labs(title = "HR1"),
                        data %>% filter(z_HR1 < 2.5) %>% ggplot(aes(x = time, y = HR1)) + geom_line() + labs(title = "HR1 corrected"),
                        data %>% ggplot(aes(x = time, y = HR2)) + geom_line() + labs(title = "HR2"),
                        data %>% filter(z_HR2 < 2.5) %>% ggplot(aes(x = time, y = HR2)) + geom_line() + labs(title = "HR2 corrected"))

#Respiration plots
gridExtra::grid.arrange(data %>% ggplot(aes(x = time, y = Resp1)) + geom_line() + labs(title = "Resp1"),
                        data %>% filter(z_RE1 < 2.5) %>% ggplot(aes(x = time, y = Resp1)) + geom_line() + labs(title = "Resp1 corrected"),
                        data %>% ggplot(aes(x = time, y = Resp2)) + geom_line() + labs(title = "Resp2"),
                        data %>% filter(z_RE2 < 2.5) %>% ggplot(aes(x = time, y = Resp2)) + geom_line() + labs(title = "Resp2 corrected"))


## Scale
### Tip: if scale() gives some issues, try the one below
z_scale <- function(column){
  column_c <- (column - mean(column)) / sd(column)
}
# Plot again to check how scaled data looks like ##############

## Now add the group, trial, condition to the cleaned up, scaled, downsampled data
## Tip the info is in the file name (GUSTAV)
```


## Now we are ready to go to load and pre-process all files

Go through all the files (with a function passed onto map_df), check which files should be excluded, if any, and save the pre-processed time-series. This procedure is similar to what you have done in portfolio 3. You may use the code you wrote for that assignment and adjust it to this one.

A couple of tips:
- looping will be too slow for these files (remember you have ~200 000 rows in each file!). Making a function and using Map/Map_df is your salvation.
- you may want your first step after loading a file to be downsampling, so that you don't work with enormous amount of data
- each study restarts the group numbering, so you should make sure to change that (e.g. 100 * Study + Group)
- you need to make sure all the data are meaningful or something has to be removed. Plotting is your friend. E.g. "Study1_G1_T1_Synchronous" has one bad respiration signal. We could replace it with NAs


```{r}
# Define a function running the loading, artifact removal, scaling, downsampling, info adding.
# Now we want to make a function which can get all the datafiles into a single dataset 
read_data <- function(filename) {
    # load data
    filename <- paste("data/", filename, sep = "")
    data <- read_csv(filename)
    # parse filename to extract study, diagnosis, subject and trial
    ss <- str_split(filename, '_')
    study <- sapply(ss, '[', 1) %>% gsub('data/Study', '', .)
    pair <- as.numeric(sapply(ss, '[', 2) %>% gsub('G', '', .))
    trial <- sapply(ss, '[', 3) %>% gsub('T', '', .)
    condition <- sapply(ss,'[', 4) %>% gsub('.csv', '', .)
    # 
    resp_1 <- data$Resp1
    resp_2 <- data$Resp2
    ecg_1 <- data$ECG1
    ecg_2 <- data$ECG2
    hr_1 <- data$HR1
    hr_2 <- data$HR2
    row_number <- seq.int(nrow(data))
    #change participant numbers for study 3 and 4 
    if (study == 3) {
    pair <- pair + 10 
    } 
    else if (study == 4) {
    pair <- pair + 20 
    }
    # combine all this data in one dataset
    return(tibble(study, pair, trial, condition, row_number, resp_1, resp_2, ecg_1, ecg_2, hr_1, hr_2))
}



# Identify all files to be read
# Run the function on the whole dataset using map_df
df <- list.files(path = "data/", pattern = ".csv") %>% ## NB replace with your path to the files
    purrr::map_df(read_data)

#removing all hr data that is below 35 and above 190 as it is unrealistic
data1 <- df %>% filter(hr_1 > 35 & hr_1 < 190,
                       hr_2 > 35 & hr_2 < 190,
                       condition != "MovementCoop", 
                       condition != "MovementGuided")


#mean centering for each individual and for each condition
data1 <- data1 %>% 
    group_by(pair, condition) %>% 
    mutate(hr_1 = hr_1 - mean(hr_1),
           hr_2 = hr_2 - mean(hr_2)
           )

str(df)

# Downsample: Since we have so much data, we want to downsample 
data1 = data1 %>%
  group(n = 100, method = 'greedy') %>%
  dplyr::summarise(
    trial = trial[1],
    row_number = row_number[1], #not an ideal solution....     
    study = study[1],
    pair = pair[1],
    condition = condition[1],
    hr_1 = mean(hr_1, na.rm = T),
    hr_2 = mean(hr_2, na.rm = T),
    resp_1 = mean(resp_1, na.rm = T),
    resp_2 = mean(resp_2, na.rm = T),
    ecg_1 = mean(ecg_1, na.rm = T),
    ecg_2 = mean(ecg_2, na.rm = T),
    )
       
# Now we need to make sure all the data are meaningful or something has to be removed
# E.g. "Study1_G1_T1_Synchronous" has one bad respiration signal. We could replace it with NAs


# plots plots plots


str(data1)
data1$pair <- as.factor(data1$pair)
data1$condition <- as.factor(data1$condition)
data1$study <- as.factor(data1$study)


####### Remove bad data
# Altering values 2.5 or more sd's from the mean to 2.5 sd's from the mean.
removeOuts <- function(ts,threshold){
  ts[ts > (mean(ts, na.rm = T) +
             (threshold * sd(ts, na.rm = T))) | 
       ts < (mean(ts, na.rm = T) -
             (threshold * sd(ts, na.rm = T)))] = mean(ts, na.rm = T)
  return(ts)
}
threshold = 2.5

data1 <- data1 %>%
  mutate(hr_1 = removeOuts(hr_1, threshold),
         hr_2 = removeOuts(hr_2, threshold), 
         resp_1 = removeOuts(resp_1, threshold),
         resp_2 = removeOuts(resp_2, threshold),
         ecg_1 = removeOuts(ecg_1, threshold),
         ecg_2 = removeOuts(ecg_2, threshold)
         )

# Save the data
write.csv(data1, "data_downs.csv") #(Tobias)
```

## Now we need to run some analysis

Let's start with a multilevel model that accounts for 
- stability (how each signal is autocorrelated)
- interpersonal dependence (each signal is dependent from the previous state of the other signal)


The data needs to be further prepared, so we can analyze both participants in the same model.
We need to turn the data into a long format:
- a column indicating own hr and ones own respiration (Y) 
- a column indicating other hr and ones other respiration (Y)
- a column indicating change in hr from previous round and one in respiration


We can then run an analysis where change is a function of one's previous state (stability, see slides), and the other's previous state (coupling). Make sure to:
- set up the most interesting contrasts: how do these parameters vary by condition? which condition should be baseline?
- set up the right random effects.
- N.B. the model will be slow. Make sure it works on a subset of the data first!

Bonus question: what if we include an additional layer? Is my heart rate just adjusting to yours, or also to how much you are adjusting to mine?
- to start answering this we can add a column indicating the previous change in hr in the other and one in respiration
- we can then build on the previous models by also adding the previous change in the other

```{r}
data2 <- read.csv("data_downs.csv")
#making a smaller dataframe for testing the functions and such
#pacman::p_load(caTools)
#set.seed(123) #randomly chosen but allows me to recreate the specific test and training set given the code "123"
#split = sample.split(data1$pair, SplitRatio = 0.9)
#data12 = subset(data1, split == FALSE)
data2$condition <- as.character(data2$condition)

data2$condition[data2$condition == "SelfPaced"] <- "Conversation" #(Morten)
```

```{r}
data2 <- data2 %>% mutate(
  # Make hr self and hr other for both participants
  hrself_1 = hr_1,
  hrother_1 = hr_2,
  hrself_2 = hr_2,
  hrother_2 = hr_1,
  # Generate a column for each: previous HR1, HR2, Resp1, Resp2. Tip: use the function Lag()
  hrlag_1 = lag(hr_1, k = 1),
  hrlag_2 = lag(hr_2, k = 1),
  resplag_1 = lag(resp_1, k = 1),
  resplag_2 = lag(resp_2, k = 1),
  ecglag_1 = lag(ecg_1, k = 1),
  ecglag_2 = lag(ecg_2, k = 1),
  # Make hr lag self and hr lag other for both participants
  hrlagself_1 = hrlag_1,
  hrlagother_1 = hrlag_2,
  hrlagself_2 = hrlag_2,
  hrlagother_2 = hrlag_1,
  # Generate a column for each: change in HR1, HR2, Resp1, Resp2 
  hrcha_1 = hr_1 - hrlag_1,
  hrcha_2 = hr_2 - hrlag_2,
  respcha_1 = resp_1 - resplag_1,
  respcha_2 = resp_2 - resplag_2,
  ecgcha_1 = ecg_1 - ecglag_1,
  ecgcha_2 = ecg_2 - ecglag_2
)

data2$hr_1 <- NULL
data2$hr_2 <- NULL
data2$hrlag_1 <- NULL
data2$hrlag_2 <- NULL #(Nikoline)
```


```{r}
# Make the data long, so we can analyze both participants at the same time. Tip: you can use the function gather ()
# Use pivot_longer to transform from wide to long format
data3 <- data2 %>% pivot_longer(
  # Select columns between resp_1 and ecgcha_2
  cols = resp_1:ecgcha_2,
  # Columns will be named '.value' and participant, .value makes one column for each unique value
  names_to = c(".value", "participant"),
  # define pattern for creating above columns, everything before the underscore goes in the value columns and everything after goes in the participant column
  names_pattern = "(.+)_(.+)"
)
data3 <- data3 %>% mutate(
  hrlagdiff = hrlagself - hrlagother
)

#as.numeric so we can have a unique id column
data3$participant <- as.numeric(data3$participant)
data3$pair <- as.numeric(data3$pair)

data3 <- data3 %>% mutate(
  participant = (pair * 10) + participant
)

data3$pair <- as.factor(data3$pair)
data3$condition <- as.factor(data3$condition)
data3$study <- as.factor(data3$study)
data3$participant <- as.factor(data3$participant)

data3$X.2 <- NULL
data3$X.1 <- NULL
data3$X <- NULL
data3$.groups <- NULL

write.csv(data3, "data.csv") #(Magnus)
```

```{r}
data <- read.csv("data.csv")
remove(data2, data3)

data$condition <- as.factor(data$condition)
data$pair <- as.factor(data$pair)
data$condition <- as.factor(data$condition)
data$study <- as.factor(data$study)
str(data)

## N.B. This is a bit tricky and you might have to do it in several steps
# model change as a function of own and other previous state
model <- lmerTest::lmer(hrcha ~ (hrlagself + hrlagdiff) * condition + ( 1 | pair) + ( 1 | study), data = data, REML = F)
summary(model)

# Set the most interesting contrast e.g. by defining synchronous or conversation as the baseline. (we are setting the same baseline once again just because we were told to)
data$condition <- relevel(as.factor(data$condition), "Conversation") #Synchronous is chosen as the baseline because it is the most basic condition. Because people speak at the same time. 
unique(data$condition)

# model change as a function of own and other previous state
model2 <- lmerTest::lmer(hrcha ~ (hrlagself + hrlagdiff) * condition + ( 1 | pair) + ( 1 | study), data = data, REML = F)
summary(model2) #excact same output as before #(Gustav)
```


## Now we need to create control baselines.

First shuffled controls, then surrogate pairs.

### Creating controls: shuffled controls

Shuffled controls break the temporal dependencies of time-series by shuffling the value within one time-series. This ensures the "coordination" observed is not due to the actual values in the series and not their sequence.
Tip: sample() is your friend, but make sure to shuffle things within participant/condition and not throughout the whole dataset
 
```{r}
# Create a shuffled dataset
d_shuffled <- data %>% group_by(pair, condition) %>% 
  mutate(
    hrlagself = sample(hrlagself),
    hrlagdiff = sample(hrlagdiff),
    hrcha = sample(hrcha),
    type = "shuffled"
)

# Concatenate it to the original dataset (and remember to have a column telling you which is which)
d_shuffled <- as.data.frame(d_shuffled)
data$type <- "original"
data_rbind <- rbind(d_shuffled, data)


# Create the same models as in the previous chunk, but adding an interaction by shuffled vs. real
data_rbind$type <- as.factor(data_rbind$type)

str(data_rbind)

data_rbind$X <- NULL

model3 <- lmerTest::lmer(hrcha ~ ((hrlagself + hrlagdiff) * condition) : type + ( 1 | pair), data = data_rbind, REML = F)
summary(model3) #(Magnus)
```
 #here we look at 
 
### TRICKY! Creating controls: surrogate pair controls
 - Per each real pair, identify at least one surrogate pair (matching one of the participants, with somebody doing the same task, but in a different pair)

```{r}
data1 <- read.csv("data_downs.csv")
remove(d_shuffled, data_rbind, model, model2, model3)

data1$condition <- as.character(data1$condition)

data1$condition[data$condition == "SelfPaced"] <- "Conversation"

# Function for making surrogate pair data
surr <- function(c, d){
  surrogate_df <- data.frame()
  study <- data1 %>% filter(study == c, condition == d)
  # Identify unique pairs within a given study (to keep things manageable) and create list of possible surrogate pairs (e.g. individual 1 from pair 1 and individual 2     from
  pairlist <- expand.grid(a = unique(study$pair), b = unique(study$pair))
  pairlist <- pairlist %>% subset(a != b)
  pair_1 <- as.vector(pairlist$a)
  pair_2 <- as.vector(pairlist$b)
  # Loop through length of pairlist
  for (i in 1:nrow(pairlist)) {
    # Make new dataframe from only current pair in loop
    subset_hr1 <- study %>% select(pair, hr_1) %>% filter(pair == pair_1[i])
    subset_hr2 <- study %>% select(pair, hr_2) %>% filter(pair == pair_2[i])
    # Make column of column number
    subset_hr1$X1 <- seq.int(nrow(subset_hr1))
    subset_hr2$X1 <- seq.int(nrow(subset_hr2))
    # If length of one is longer than other, change length of longer one to length of shorter one
    subset_hr1 <- subset_hr1 %>% filter(X1 <= length(subset_hr2$X1))
    subset_hr2 <- subset_hr2 %>% filter(X1 <= length(subset_hr1$X1))
    # Rename
    subset_hr1 <- rename(subset_hr1, "pair_hr1" = "pair")
    subset_hr2 <- rename(subset_hr2, "pair_hr2" = "pair")
    # Join hr 1 and hr 2
    cbind_df <- cbind(subset_hr1, subset_hr2)
    # Remove column numbers
    cbind_df$X1 <- NULL
    cbind_df$X1 <- NULL
    # Define condition and study
    cbind_df$condition <- d
    cbind_df$study <- c
    # add to dataframe
    surrogate_df = rbind(surrogate_df, cbind_df)
  }
  return(surrogate_df)
}

a <- surr(1, "Conversation")
b <- surr(1, "Synchronous")
c <- surr(1, "TurnTaking")

abc <- rbind(a, b, c)
abc <- abc %>% subset(pair_hr1 != 900)

abc2 <- abc %>% mutate(
  hrlag_1 = lag(hr_1, k = 1),
  hrlag_2 = lag(hr_2, k = 1),
  # Generate a column for each: change in HR1, HR2, Resp1, Resp2 
  hrcha_1 = hr_1 - hrlag_1,
  hrcha_2 = hr_2 - hrlag_2,
)

abc <- abc %>% mutate(
  # Make hr self and hr other for both participants
  hrself_1 = hr_1,
  hrother_1 = hr_2,
  hrself_2 = hr_2,
  hrother_2 = hr_1,
  # Generate a column for each: previous HR1, HR2, Resp1, Resp2. Tip: use the function Lag()
  hrlag_1 = lag(hr_1, k = 1),
  hrlag_2 = lag(hr_2, k = 1),
  # Make hr lag self and hr lag other for both participants
  hrlagself_1 = hrlag_1,
  hrlagother_1 = hrlag_2,
  hrlagself_2 = hrlag_2,
  hrlagother_2 = hrlag_1,
  # Generate a column for each: change in HR1, HR2, Resp1, Resp2 
  hrchaself_1 = hr_1 - hrlag_1,
  hrchaself_2 = hr_2 - hrlag_2,
  # Make new pair column
  pair = pair_hr1 * 10 + pair_hr2
  )

# Make the data long, so we can analyze both participants at the same time. Tip: you can use the function gather ()
# Use pivot_longer to transform from wide to long format
abc_long <- abc %>% pivot_longer(
  # Select columns between resp_1 and ecgcha_2
  cols = hrself_1:hrchaself_2,
  # Columns will be named '.value' and participant, .value makes one column for each unique value
  names_to = c(".value", "participant"),
  # Define pattern for creating above columns, everything before the underscore goes in the value columns and everything after goes in the participant column
  names_pattern = "(.+)_(.+)"
)

# Create column hr lag diff
abc_long <- abc_long %>% mutate(
  hrlagdiff = hrlagself - hrlagother
)

# Relevel 
abc_long$condition <- relevel(as.factor(abc_long$condition), "Conversation") #Synchronous is chosen as the baseline because it is the most basic condition. Because 

model_4 <- lmerTest::lmer(hrchaself ~ (hrlagself + hrlagdiff) * condition + ( 1 | pair), data = abc_long, REML = F)
summary(model_4) #(Morten)

```
 

### Effects of respiration coordination on heart rate coordination
 - describe how you would test those.
 - Optional: run the models and report them
